%*******************************************************
% Abstract
%*******************************************************
%\renewcommand{\abstractname}{Abstract}
\pdfbookmark[1]{Abstract}{Abstract}
\begingroup
\let\clearpage\relax
\let\cleardoublepage\relax
\let\cleardoublepage\relax

\chapter*{Abstract}
Powerful modern automated theorem provers (ATP) are increasingly used to establish the correctness of critical computer systems.
Among them, Satisfiability Modulo Theories (SMT) solvers play a central role thanks to their expressive power and efficiency in handling complex logical reasoning tasks.
However, their size and complexity make full formal verification of their implementations impractical, raising a fundamental question of trust:
\begin{center}
\emph{Can we trust the correctness of their results?}
\end{center}

A promising answer lies in \emph{proof logging}, where solvers produce proof traces that can be independently checked by simpler and verifiable proof checkers.
The Alethe format has recently emerged as a unified framework for representing such SMT proofs.
The Alethe format has been adopted by several SMT solvers for explaining why a set of constraints is unsatisfiable.
Yet, Alethe lacks a certified proof checker, limiting its adoption and trustworthiness.

In this thesis, we propose a reconstruction framework that translates Alethe proof traces into the Lambdapi proof assistant, an implementation of the $\lpm$-calculus modulo rewriting.
Lambdapi is a foundational proof assistant based on dependent type theory and rewriting rules, intended to serve as a pivot for exchanging proofs between interactive proof assistants.
We present a modular encoding of SMT logic and Alethe rules in Lambdapi, covering core theories such as propositional reasoning and linear arithmetic.

We detail the reconstruction process, including the elaboration of solver-specific simplifications, the reflective normalization of arithmetic expressions, and the handling of subproofs and quantified formulas.
Finally, we evaluate our implementation on benchmark examples and demonstrate that Alethe proofs generated by modern SMT solvers can be efficiently and reliably reconstructed in Lambdapi, paving the way toward portable and independently verifiable SMT proofs.

\endgroup			

\vfill